---
title: "Working with Web API Data in R"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Working with APIs in R
Making HTTP requests to APIs from R can be really simple. The `httr` and `jsonlite` libraries are probably the most common tools for this process, so let's install and import those:

```{r}
install.packages(c("httr", "jsonlite"))
library(httr)
library(jsonlite)
library(tidyverse)
```

## Authentication
You'll have to authenticate each request to the Harvard Art Museum API with an API key. Other APIs may require different kinds of authentication (sometimes very complicated auth! Look for libraries at that point), but HAM has some pretty simple authentication, which makes things easy for us.

You can sign up for a key [here](https://www.harvardartmuseums.org/collections/api). Documentation for the entire API is hosted on GitHub and can be viewed [here](https://github.com/harvardartmuseums/api-docs). It's quite a comprehensive series of docs, so you should check it out.

Recently, the Art Museum also added an [API Tookit and Guide](https://api-toolkit.herokuapp.com/). This is another fantastic resource.

Let's store our API key in a variable so we can reuse it easily:
```{r}
APIKEY <- ""
```

## Basic Requests

We're going to start off with a basic request to the API. This API, like many others, has a variety of endpoints, each with their own url, slightly modified from a base url. We'll worry about the general case in a bit, for now let's look at a basic API request.

In this example, we'll re-create the first example in the [Object endpoint documentation](https://github.com/harvardartmuseums/api-docs/blob/master/sections/object.md), which will give each of you the records for 10 objects that have never been viewed online in the museum's collections.

```{r}
# httr::set_config(httr::config(ssl_verifypeer = 0L)) 
endpoint <- "https://api.harvardartmuseums.org/object"
parameters <- list(
  q = "totalpageviews:0",
  size = 10,
  apikey = APIKEY,
  sort = "random"
)
request <- GET(url = endpoint, query = parameters)
```

We now have a request object, which includes a lot of data. Let's take a look:
```{r}
request
```
This shows us
- the URL which `httr` created from the endpoint and the query parameters we provided
- the date we received the response
- the status of the response, which is hopefully 200. See more about [HTTP response codes](https://developer.mozilla.org/en-US/docs/Web/HTTP/Status)
- the content type of the response - JSON
- and the size of the response

We can check out the type of the object:
```{r}
typeof(request)
```

As a refresher: lists in R can have named elements. We can use that name as a key to access particular elements. There are also special methods for accessing some kinds of content, such as the HTTP status code, the headers, and the body content.
```{r}
print(paste("URL: ", request$url))
print(paste("Status: ", status_code(request)))
print("Headers:")
headers(request)
```

We're interested in using the response data returned from our request. Let's check that out:
```{r}
response <- content(request, as = "text", encoding = "UTF-8")
response
```

That's the data we want, but it's not going to be very easy to use in this format. Let's use `jsonlite` to turn it into a dataframe.

```{r}
df <- fromJSON(response, flatten = TRUE) %>% 
  data.frame()
df
```

Now we have a dataframe, which is the preferred format to use within R for most applications. Let's package up this transformation snippet into a function.

```{r}
request_to_dataframe <- function(req){
  res <- content(req, as = "text", encoding = "UTF-8")
  df <- fromJSON(res, flatten = TRUE) %>% 
    data.frame()
  return(df)
}
```

## Changing our request

Let's say we're not interested in the most obscure parts of the collection, but rather in the most popular parts of the collection. There are a few ways we might go about doing this. One way might be to sort our search results by `totalpageviews`, and see what the top 10 are.

To do that, we can go back to the [Object API documentation](https://github.com/harvardartmuseums/api-docs/blob/master/sections/object.md) and look for hints about what we might be able to do.

```{r}
params <- list(
  apikey = APIKEY,
) # ADD YOUR QUERY HERE
request <- GET(url = endpoint, query = params)
top_items_df <- request_to_dataframe(request)
top_items_df
```

### Looking at our Results
That's a lot of info. Let's try to get just the title, date, pageview count, and URL for each record.
```{r}
top_items_small <- top_items_df %>% select("records.title", "records.dated", "records.totalpageviews", "records.url")
top_items_small
```

The top result from this query is a Van Gogh painted titled "Self-Portrait Dedicated to Paul Gauguin." You can grab just the first object with the head() method, which returns the first row in the dataset, or by accessing the records list (which is indexed from 0). Let's take a look at this most popular painting.
```{r}
# topResult = top_items_df[1,] is another way of doing this - it says "get a sublist from the first index"
head(topResult)
topResult
```

### Exercise
- Try using the `person` endpoint to search for information about Van Gogh. Get his `id` number.
- Try displaying all HAM works by Van Gogh using that `id`. Filter your results to only include records with an image associated.

```{r}
# Your code goes here
# You will need to change the endpoint we've been using
```

### Endpoint Exercises
[HAM API Documentation](https://github.com/harvardartmuseums/api-docs)

#### Get a list of all the medium types in the museum
```{r}
# Your code goes here

```

#### Find out all the exhibits currently being shown at the Harvard Art Museums venue
```{r}
# Your code goes here

```

#### Given the above, create a dataframe of all the objects currently being shown in these exhibits
- Your dataframe should only have some fields: title, dated, total page views, accession year, classification, technique, medium, images
- First, create a function which will get all of the objects from one exhibit. The function will need to page through the exhibit if there are multiple pages
- Then, apply the function to all the exhibits
- Return the dataframe from the function
```{r}
get_exhibit_items <- function(exhibit_id){
  params <- list(
    apikey = APIKEY,
    # your query goes here
  )
  endpoint  = "https://api.harvardartmuseums.org/object"
  init_req <- GET(url = endpoint, query = params)
  init_req
  
  # turn it into a dataframe
  df <- request_to_dataframe(init_req)
  
  # get rid of any column which starts with .info
  
  # parse the init_req
  
  # get the next_link from the parsed init_req
  
  # if the next_link exists, loop through all of the other pages until it's gone
  # in each iteration, parse the content and bind it to the existing dataframe (df) so it persists outside the loop
  # get the next loop and continue
  # hint: a while loop might be useful here
  
  # now return our dataframe df
  
}

all_items = data.frame()
exhibit_ids_list <- as.list(exhibits_df$records.id)

for(id in exhibit_ids_list){
  print(id)
  items <- get_exhibit_items(id)
  print(items)
  all_items <- bind_rows(all_items, items)
}
all_items

```

## IIIF
When we printed the 10 most popular records above (under **Looking at our Results**), you may have noticed a sharp dropoff after the first few records. Our Van Gogh painting is particularly popular, with ~10000 more views than the second most popular record and 4x as many as the tenth most popular. This particular Art Museum record is used as the default image asset for the demo installation of [Project Mirador](http://projectmirador.org/demo/), an image viewer for [IIIF (International Image Interoperability Framework)](https://iiif.io/) media assets. 

We're not going to go deep into IIIF in this workshop, but want to mention that IIIF is both a community of developers and a collection of APIs and API-compliant tools that you can use to share, manipulate, and display visual materials. The [Image API](https://iiif.io/api/image/2.1/) and [Presentation API](https://iiif.io/api/presentation/2.1/) are the most used outputs as of now, though there are also APIs for Authentication, Search, and beta versions for other media (video and VR).

Let's return to our top result, Van Gogh's "Self-Portrait Dedicated to Paul Gauguin", and look specifically at the `records.images` column:
```{r}
topResult$records.images
```
This particular record has 6 images associated with it. Try copying and pasting some of the `baseimageurl`s in your browser.

You'll notice that the `baseimageurls` use Harvard's Name Resolution service, which redirects to an Image Delivery Service URL that displays the image. This is nice, but we're more interested in the iiifbaseuris because we can manipulate IIIF resources using the Image API. Try opening one of those. What happens?

The IIIF Image API spec requires that we pass not just a baseurl, but a well-formed IIIF-compliant URI to get an image. Let's check out that [documentation](https://iiif.io/api/image/2.1/) and see what else we need to construct one of those.

From the docs:

>The IIIF Image API URI for requesting an image must conform to the following URI Template:
>
>`{scheme}://{server}{/prefix}/{identifier}/{region}/{size}/{rotation}/{quality}.{format}`
>
>For example:
>
>`http://www.example.org/image-service/abcd1234/full/full/0/default.jpg`
>
The parameters of the Image Request URI include region, size, rotation, quality and format, which define the characteristics of the returned image. These are described in detail in Image Request Parameters.

The `iiifbaseuri`s include up through the `{identifier}`, but we need to include additional parameters to get the server to actually render the image for us. These parameters are passed within the URI itself, rather than in a query string appended after a delimiter (usually `?`), which is what we've been using `httr` to do. Let's write a function that can generate IIIF URIs for us.

```{r}
ham_iiif_query <- function(baseuri, region="full", size="full", rotation=0, quality="default", format="jpg"){
  if(substr(baseuri, nchar(baseuri) - 2, nchar(baseuri)) != "/"){
    baseuri <- paste(baseuri, "/", sep="")
  }
  url <- str_interp("${baseuri}${region}/${size}/${rotation}/${quality}.${format}")
  return(url)
}
```

```{r}
ham_iiif_query("https://ids.lib.harvard.edu/ids/iiif/47174896")
```

You should be able to open that link in your browser. Let's try to use this function to display the images within our notebook:
```{r}
#install.packages("imager")
#library("imager")
van_gogh_images <- topResult$records.images[[1]]
typeof(van_gogh_images)
length(van_gogh_images)

van_gogh_image_uris <- as.list(van_gogh_images$iiifbaseuri)
for(iiif_uri in van_gogh_image_uris){
  url <- ham_iiif_query(iiif_uri)
  print(url)
  img <- load.image(url)
  plot(img)
}
```

### IIIF Image API Exercise
We can easily manipulate the images to via the IIIF image API. Let's try requesting only the right half of an image (using `region`), in black and white, scaled to 300px wide, and getting back a PNG:
```{r}
for(iiif_uri in van_gogh_image_uris){
  # your code goes here
}
```
### IIIF Presentation API

If you're interested in the Presi API (for presenting structured IIIF resources as part of a more fully-functional web app), check out [this documentation](https://iiif.io/api/presentation/2.1/) to learn how IIIF manifests structure sequences of canvases which image viewers then present to end users. You can find an HAM Object's manifest in the `seeAlso` field, or by by appending the object ID to a baseurl:
```{r}
topResult$records.seeAlso
str_interp("https://iiif.harvardartmuseums.org/manifests/object/${topResult['records.id']}")
```

#### Mirador

You can consume these resources using [Mirador](http://projectmirador.org/), an image viewer which uses the IIIF Image and Presentation APIs.

If you head to the [Project Mirador Demo](http://projectmirador.org/) page, you can add a new manifest in the top left ("four boxes icon" -> "Replace Object" -> "Add new object from URL"). You may notice that the Van Gogh is one of the example objects here, so let's instead add a manifest from our second most viewed painting, "The Gare Saint-Lazare: Arrival of a Train": https://iiif.harvardartmuseums.org/manifests/object/228649